Liveness-Based Garbage Collection for Lazy Languages

### 2016-04-26 paper/paper.tex
Fixed comments except WHNF

While  the runtime  system  of most  functional  languages includes  a
garbage collector to efficiently  reclaim memory, empirical studies on
Scheme~\cite{karkare06effectiveness}   and,   more   importantly,   on
Haskell~\cite{rojemo96lag} programs have shown that garbage collectors
leave uncollected a large number  of memory objects that are reachable
but not live (here {\em live} means the object can potentially be used
by the program at a  later stage).  This results in unnecessary memory
retention.

@@

While  the idea  of  using static  analysis  to improve  memory
  utilization  has  been  shown  to   be  effective  for  {\em  eager}
  languages~\cite{asati14lgc,       HofmannJ03,       inoue88analysis,
    lee05static}, a straightforward extension  of the technique is not
  possible  for  lazy  languages,  where  heap-allocated  objects  may
  include   closures    (runtime   representations    of   unevaluated
  expressions).
%%\todelete{Designing  a  liveness  based  garbage
%%  collector   for  lazy   languages  poses   significant  challenges.}
Firstly, since  data is made  live by  evaluation of closures,  and in
lazy languages  the place in  the program where this  evaluation takes
place cannot  be statically determined, laziness  complicates liveness
analysis itself.  Moreover, for liveness-based  GC to be effective, we
need to  extend it  to closures  apart from  evaluated data.   Since a
closure can escape the scope in which  it was created, it has to carry
the  liveness  information  of   its  free  variables.   As  execution
progresses  and possible  future uses  are eliminated,  we update  the
liveness  information  in  a  closure with  a  more  precise  version.
Because  of   these  reasons,  the  garbage   collector  also  becomes
significantly more  complicated than a liveness-based  collector for a
eager language.

@@

%------------------------------------------------------------%
\subsection{Motivating example}
\label{sec:motiv}
\input{fig_motiv}   
%%
Figure~\ref{fig:mot-example}   shows  an   example
program and the state of the heap at the program point $\pi$\footnote{We write
$\pi\!:\!e$ to  associate a  label $\pi$ with  the program  point just
before  the  expression  $e$. The label is not part of the syntax.},  i.e., just  before  the  evaluation  of
$(\length\ \pz)$.  The heap is represented  by a graph in which a node
either  represents  atomic  values   ($\NIL$,  integers  etc.),  or  a
\CONS\  cell  containing  $\CAR$  and  $\CDR$  fields,  or  a  closure
(represented  by  shaded  clouds).   Edges   in  the  graph  are  {\em
  references} and represent values of variables or fields.  The figure
shows the  lists \px\  and \pz\  partially evaluated  due to  the
\SIF\ condition 
(\NULLQ~(\CAR~\pz)).
The edges  shown by thick  arrows are those  which are live  at $\pi$.
%% Only those  cell which have a future use  should be preserved
%% during garbage collection, all other
%% cells can be  reclaimed.
Thus if a GC takes  place at $\pi$ with the heap shown
in Figure~\ref{fig:mot-example}(b),  a liveness-based  collector (LGC)
will preserve only  the cells referenced by $\pz$, and  the live cells
constituting the  closure referenced by $(\CDR~\pz)$.   In contrast, a
reachability-based collector (RGC) will preserve all cells.

@@

\section{Towards a  computable form of liveness}\label{sec:computing}
%\section{Generating and storing liveness information}\label{sec:computing}
The analysis in Section~\ref{sec:liveness} is fully context-sensitive, 
describing the  liveness sets  in a function body in
terms of a symbolic demand $\sigma$  and \Lfonly. However, we are yet to
describe (i) how to obtain demand transformers \Lfonly\ from the rule
{\sc live-define} , and (ii) how to compute the concrete demand $\sigma$ on
each function. To do so, we first need to modify the liveness rules
to a slightly different form.

@@

In addition to  evaluated data, our collector  also reclaims closures.
For this, we  had to modify the standard run-time  representation of a
closures  to carry  liveness  of its free  variables and  periodically
update the liveness during execution  time.  As expected, our garbage
collector  reclaims more  garbage and  reduces memory  requirement of
programs.  An  additional benefit is  that in  spite of using  a more
expensive collector, the execution times remains comparable in
most cases, and even improves for some programs.

@@

  Orthogonally,   we   plan  to   improve   the   efficiency  of   the
  liveness-based garbage  collector using heuristics such  as limiting
  the depth of DFA, merging  nearly-equivalent states and using better
  representation(for example BDDs~\cite{Bryant86})  and algorithms for
  automata manipulation.  We also  need to investigate the interaction
  of liveness with  other collection schemes, such  as incremental and
  generational collection.   It might  be interesting  to use  a mixed
  mode GC  scheme which allows the  costs of LGC to  be amortized over
  several runs of RGC. In summary, we need to investigate ways to make
  liveness-based GC attractive for practical collectors.

>>>>

While  the runtime  system  of most  functional  languages includes  a
garbage collector to efficiently  reclaim memory, empirical studies on
Scheme~\cite{karkare06effectiveness}  and,  Haskell~\cite{rojemo96lag}
programs have shown that garbage  collectors leave uncollected a large
number of  memory objects that are  reachable but not live  (here {\em
  live} means the  object can potentially be used by  the program at a
later stage).  This results in unnecessary memory retention.

@@

While  the idea  of  using static  analysis  to improve  memory
  utilization  has  been  shown  to   be  effective  for  {\em  eager}
  languages~\cite{asati14lgc,       HofmannJ03,       inoue88analysis,
    lee05static}, a straightforward extension  of the technique is not
  possible  for  lazy  languages,  where  heap-allocated  objects  may
  include   closures    (runtime   representations    of   unevaluated
  expressions).
%%\todelete{Designing  a  liveness  based  garbage
%%  collector   for  lazy   languages  poses   significant  challenges.}
Firstly, since  data is made  live by  evaluation of closures,  and in
lazy languages  the place in  the program where this  evaluation takes
place cannot  be statically determined, laziness  complicates liveness
analysis itself.  Moreover, for liveness-based  GC to be effective, we
need to  extend it  to closures  apart from  evaluated data.   Since a
closure can escape the scope in which  it was created, it has to carry
the  liveness  information  of   its  free  variables.   As  execution
progresses  and possible  future uses  are eliminated,  we update  the
liveness  information  in  a  closure with  a  more  precise  version.
For these  reasons,  the  garbage   collector  also  becomes
significantly more  complicated than a liveness-based  collector for an
eager language.

@@

%------------------------------------------------------------%
\subsection{Motivating example}
\label{sec:motiv}
\input{fig_motiv}   
%%
Figure~\ref{fig:mot-example}   shows  an   example
program and the state of the heap at the program point $\pi$\footnote{We write
$\pi\!:\!e$ to  associate a  label $\pi$ with  the program  point just
before  the  expression  $e$. The label is not part of the syntax.},  i.e., just  before  the  evaluation  of
$(\length\ \pz)$.  The heap is represented  by a graph in which a node
either  represents  atomic  values   ($\NIL$,  integers,  etc.),  or  a
\CONS\  cell  containing  $\CAR$  and  $\CDR$  fields,  or  a  closure
(represented  by  shaded  clouds).   Edges   in  the  graph  are  {\em
  references} and represent values of variables or fields.  The figure
shows the  lists \px\  and \pz\  partially evaluated  due to  the
\SIF\ condition 
(\NULLQ~(\CAR~\pz)).
The edges  shown by thick  arrows are those  which are live  at $\pi$.
%% Only those  cell which have a future use  should be preserved
%% during garbage collection, all other
%% cells can be  reclaimed.
Thus if a GC takes  place at $\pi$ with the heap shown
in Figure~\ref{fig:mot-example}(b),  a liveness-based  collector (LGC)
will preserve only  the cells referenced by $\pz$, and  the live cells
constituting the  closure referenced by $(\CDR~\pz)$.   In contrast, a
reachability-based collector (RGC) will preserve all cells.

@@

\section{Towards a  computable form of liveness}\label{sec:computing}
%\section{Generating and storing liveness information}\label{sec:computing}
The analysis in Section~\ref{sec:liveness} is fully context-sensitive, 
describing the  liveness sets  in a function body in
terms of a symbolic demand $\sigma$  and \Lfonly. However, we have yet to
describe (i) how to obtain demand transformers \Lfonly\ from the rule
{\sc live-define} , and (ii) how to compute the concrete demand $\sigma$ on
each function. To do so, we first need to modify the liveness rules
to a slightly different form.

@@

In addition to  evaluated data, our collector  also reclaims closures.
For this, we  had to modify the standard run-time  representation for
closures  to carry  liveness  of its free  variables and  periodically
update the liveness during execution  time.  As expected, our garbage
collector  reclaims more  garbage and  reduces memory  requirement of
programs.  An  additional benefit is  that in  spite of using  a more
expensive collector, the execution times remains comparable in
most cases, and even improves for some programs.

@@

  Orthogonally,   we   plan  to   improve   the   efficiency  of   the
  liveness-based garbage  collector using heuristics such  as limiting
  the depth of DFA, merging  nearly-equivalent states and using better
  representation (for example BDDs~\cite{Bryant86})  and algorithms for
  automata manipulation.  We also  need to investigate the interaction
  of liveness with  other collection schemes, such  as incremental and
  generational collection.   It might  be interesting  to use  a mixed
  mode GC  scheme which allows the  costs of LGC to  be amortized over
  several runs of RGC. In summary, we need to investigate ways to make
  liveness-based GC attractive for practical collectors.
